---
title: "EDA-challenge"
author: "J.L. James-Rollins"
format: html
editor: visual
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(tidylog)
library(dplyr)
```

My method for uploading the dataset is as follows:

```{r, message=FALSE, warning=FALSE}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/data-wrangling.csv"

d <- read_csv(f, col_names = TRUE)
```

**1.** After uploading, I created a new variable called **BSD** (body size dimorphism), representing the ratio of average male body mass to average female body mass.

```{r, message=FALSE, warning=FALSE}
d <- d %>%
  mutate(BSD = Body_mass_male_mean/Body_mass_female_mean)
```

**2.** Next, I created a variable called **sex_ratio**, representing the ratio of the number of adult females to adult males in a typical group.

```{r, message=FALSE, warning=FALSE}
d <- d %>%
  mutate(sex_ratio = AdultFemale/AdultMales)
```

**3.** Following that, I added a new variable called **DI** (defensibility index), which is the ratio of day range length to the diameter of the home range.

```{r, message=FALSE, warning=FALSE}
d <- d %>%
  mutate(DI = DayLength_km/(2 * sqrt(HomeRange_km2 / pi)))
```

**4.** I plotted the relationship between *day range length* (y-axis) and *time spent moving* (x-axis) for these primate species, both overall and by family.

I found that species that spent more time moving did not travel the farthest; in fact, the species that covered the greatest distances tend to spend less time moving overall.

**Overall:**

```{r, message=FALSE, warning=FALSE}
p1 <- ggplot(data = d, aes(x = (Move), y = (HomeRange_km2),
                          color = factor(Species))) +
                            xlab("time spent moving") +
                            ylab("day range length") +
                            geom_point(na.rm = TRUE) +
                            theme(legend.position = "none")

print(p1)
```

For the different families, similar trends were observed, although the Cercopithecidae family shows some indication that time spent moving is positively correlated with the distance traveled.

**Family:**

```{r, message=FALSE, warning=FALSE}
p2 <- ggplot(data = d, aes(x = (Move), y = (HomeRange_km2),
                          color = factor(Species))) +
                            xlab("time spent moving") +
                            ylab("day range length") +
                            geom_point(na.rm = TRUE) +
                            facet_wrap(~Family, ncol = 4) +
                            theme(legend.position = "none")

print(p2)
```

Since the variables represent distance and time, there was no need to transform either of them.

**5.** I Plotted the relationship between *day range length* (y axis) and *group size* (x axis), overall and by family.

**Overall:**

```{r, message=FALSE, warning=FALSE}
p3 <- ggplot(data = d, aes(x = log(MeanGroupSize), y = (HomeRange_km2),
                           color = factor(Species))) +
                             xlab("log(group size)") +
                             ylab("day range length") +
                             geom_point(na.rm = TRUE) +
                             theme(legend.position = "none")

print(p3)
```

**Family:**

```{r, message=FALSE, warning=FALSE}
p4 <- ggplot(data = d, aes(x = log(MeanGroupSize), y = (HomeRange_km2),
                           color = factor(Species))) +
                             xlab("log(group size)") +
                             ylab("day range length") +
                             geom_point(na.rm = TRUE) +
                             facet_wrap(~Family, ncol = 4) +
                             theme(legend.position = "none")

print(p4)
```

Through the analysis, we can suggest that larger groups tend to travel farther overall, with the dominant family in this trend being Cercopithecidae.

Additionally, transforming the "group size" variable is both possible and recommended to enhance the interpretation of the data.

**6.** I plotted the relationship between *canine size dimorphism* (y axis) and *body size dimorphism* (x axis) overall and by family.

**Overall:**

```{r, message=FALSE, warning=FALSE}
p5 <- ggplot(data = d, aes(x = log(BSD), y = log(Canine_Dimorphism),
               color = factor(Species))) +
  xlab("log(body size dimorphism)") +
  ylab("log(canine size dimorphism)") +
  geom_point(na.rm = TRUE) +
  theme(legend.position = "none")

print(p5)
```

**Family:**

```{r, message=FALSE, warning=FALSE}
p6 <- ggplot(data = d, aes(x = log(BSD), y = log(Canine_Dimorphism),
               color = factor(Species))) +
  xlab("log(body size dimorphism)") +
  ylab("log(canine size dimorphism)") +
  geom_point(na.rm = TRUE) +
  facet_wrap(~Family, ncol = 4) +
  theme(legend.position = "none")

print(p6)
```

The graphs reveal that taxa with greater size dimorphism also exhibit greater canine dimorphism.

**7.** I created a variable called diet_strategy, which is labeled as “frugivore” if fruits make up more than 50% of the diet, “folivore” if leaves make up more than 50%, and “omnivore” if neither of these conditions is true and diet data is available.

```{r, message=FALSE, warning=FALSE}
d <- d %>%
  mutate(diet_strategy = case_when(
    Fruit >= 50 ~ "frugivore", 
    Leaves >= 50 ~ "folivore",
    Fruit < 50 & Leaves < 50 ~ "omnivore",  
    TRUE ~ NA
    ))
```

Before creating the boxplot, I first removed the NA values relating to the diet_strategy from the dataset.

```{r, message=FALSE, warning=FALSE}
d2 <- d %>%
  drop_na(diet_strategy)

```

Then created boxplots of group size for species with different dietary strategies.

```{r, message=FALSE, warning=FALSE}

p7 <- ggplot(data = d2, aes(x = diet_strategy, y = log(MeanGroupSize))) +
               geom_boxplot(na.rm = TRUE) +
               theme(axis.text.x = element_text(angle = 90)) +
               xlab("Diet Strategy") +
               ylab("Group Size")

print(p7)
```

The boxplot suggests that frugivores live in larger groups than folivores.

**8.** In one line of code, using {dplyr} verbs and the forward pipe (\|\> or %\>%) operator, do the following:

```         
Add a variable, Binomial to the data frame d, which is a concatenation of the Genus and Species variables…
Trim the data frame to only include the variables Binomial, Family, Brain_size_species_mean, and Body_mass_male_mean…
Group these variables by Family…
Calculate the average value for Brain_Size_Species_Mean and Body_mass_male_mean per Family (remember, you may need to specify na.rm = TRUE)…
Arrange by increasing average brain size…
And print the output to the console
```

```{r, message=FALSE, warning=FALSE}
d3 <- d %>%
  mutate(Binomial = paste(Genus, Species, sep = " ")) %>%
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) %>%
  group_by(Family) %>%
  summarise(
    Average_Brain_Size = mean(Brain_Size_Species_Mean, na.rm = TRUE),
    Average_Body_Mass = mean(Body_mass_male_mean, na.rm = TRUE)
  ) %>%
  arrange(Average_Brain_Size)

print(d3)
```
